Fields COVID-19 dataset exploration

### 2023-01-05 data_cleaning/Data_cleaning_and_covariates_Jan_05.qmd
need to write model equation, fix model and read about random effects

Below is the information provided by the 2016 Canada Census for Ontario for different visible minorities and ethnic origins. [Link to Census Data](https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/prof/details/page.cfm?Lang=E&Geo1=PR&Code1=35&Geo2=PR&Code2=01&SearchText=Ontario&SearchType=Begins&SearchPR=01&B1=Ethnic%20origin&TABID=1&type=1).

| Ethnicity/Race | percentage |
|----------------|:-----------|
| Arab           | 1.6%       |
| Black          | 4.7%       |
| East Asian     | 6.6%       |
| Indigenous     | 3.9%       |
| Latin American | 1.5%       |
| Mixed          | 1.0%       |
| Other          | 0.7%       |
| South Asian    | 8.7%       |

: Data from the 2016 Census for Ontario

In the table above, the following data were obtained from the "Visible Minorities" from the Census:

-   Arab
-   Black
-   South Asian
-   East Asian (computed by adding the Chinese, Korean and Japanese entries)
-   Mixed ("multiple visible minorities" entry in the census)
-   Other (from the "visible minority n.i.e" entry in the census)

The following data were obtained from the "ethnic origin" data from the census: - Indigenous - European Origins

However, according to the census "ethnicity" and "visible minority" seem to have different values. For example, in the Visible Minority entry of the census Latin American is 1.5%, but in the "Ethnic Origin" data the value is 2.4%.

Moreover, I would need to adjust the data for Pacific Islanders, as the survey included East Asian and Pacific Islanders in the same question but they are categorized differently in the census.

Lastly, the survey had "white caucasian" as one of the options but that category is not present in the census data. The most approximate entry on the census is "European origins" (61.6%).

**Overall, we weighted the data because it is not representative of the population distribution in Ontario**.

Next, I was interested in analyzing the geographical distribution of the participants over time, to see the percentage of answers from Toronto. This is important to see where the majority of the answers came from (is there such thing as geographical weighting?). The same will be done with the age groups of the participants, as this is data that is reported in the census.

p1<-clean_data%>%
  select(city)%>%
  mutate(location=as.character(case_when(city=="Toronto"~"Toronto",
                            city!="Toronto"~ "All other cities"
                          )))

p1$location<-as.factor(p1$location)

ggplot(p1 %>% count(location) %>%    # then count number in each group
         mutate(pct=n/sum(n),               # Calculate percent within each group
                ypos = 200),  # Calculate label positions
       aes(location, n, fill=location)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct*100),"%"), y=ypos))+
  theme_classic()+
  thm2

The proportion of responses from Toronto is 30.9% of the total. The Census data indicates that Toronto has around 38% of the total population province (5,433,590/14,223,942). This would mean that a correction would be needed also for the proportion of answers.

Additionally, the data needs to be cleaned as some of the responses do not have an assigned city, and the entry only reads "None". These entries will be removed as they lack geographical information.

```{r, remove-entries}

@@

The final aspect I wanted to explore was the income distribution. In this regard the question from the survey was "What is your household annual income?". To compare the results of this answer, I used the "Household total income groups in 2015 for private households" from the Census data. The information from the census appears next.

| Household income range (CAD) | Percentage |
|------------------------------|:-----------|
| \< 15,000                    | 5.7%       |
| 15,000 - 24,999              | 7.5%       |
| 25,000 - 39,999              | 11.6%      |
| 40,000- 59,999               | 15.4%      |
| 60,000 - 89,999              | 19.5%      |
| \>90,000                     | 40.4%      |

@@

One thing to keep in mind in this case is that the brackets for income in the census data are different than the brackets used in the survey. The census does CAD 4,999 brackets (e.g., CAD 5,000- CAD 9,9999) up to CAD 49,999 and then it does CAD 9,999 brackets up to CAD 99,999 but after that, the brackets increase to CAD 24,999, and therefore, one cannot obtain percentages for the 90,000-109,999 and \>110,000 brackets from the survey.

@@

p3<-clean_data%>%
  select(date,income)%>%
  mutate(income_ord = fct_relevel(income,
                                  "under_15000",
                                  "15000_24999",
                                  "25000_39999",
                                  "40000_59999",
                                  "60000_89999",
                                  "90000_109999",
                                  "over_110000"
                                ))

ggplot(p3%>% count(income_ord)%>%  # count number in each group
         mutate(pct=n/sum(n),               # Calculate percent within each group
                ypos = 200),  # Calculate label positions
       aes(income_ord, n, fill=income_ord)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct*100),"%"), y=ypos))+
  theme_classic()+
  thm3

@@

# Joining the municipalities data

We have obtained and cleaned the data of the municipalities from the province of Ontario website, and cleaned the dataset to obtain the city names and geographical locations. Further details can be found in `municipalities.qmd` in the `data_cleaning` directory. Next, we will join and match the geographical location from the municipalities dataset to the clean dataset we have obtained in this document so far.

@@

## Merging missing municipalities and geographic area names

@@

# Raking

The next step is to adjust the variables according to the Census data (raking). I will follow the steps provided by [https://sdaza.com/blog/2012/raking/](https://sdaza.com/blog/2012/raking/) https://www.r-bloggers.com/2018/12/survey-raking-an-illustration/

The variables to define raking weights are:

  - income
  - age groups 
  - geographical location
  - race and ethnicity: check proportions here https://www.ontario.ca/document/2016-census-highlights/fact-sheet-9-ethnic-origin-and-visible-minorities

```{r,define-target-list}

library(weights)

#get percentages
wpct(clean_data$income)
wpct(clean_data$age_group)
wpct(clean_data$race)

#create target list for weights

age_group<-c(0.127,0.129,0.128,0.149,0.137,0.167)

race<-c(0.016,0.047,0.066,)

income<-c(0.075,0.116,0.154,0.195,)

@@

>>>>

Because the categories for race/ethnicity in the survey did not match the categories from the Canadian Census, we used a combination of sources to obtain the corresponding percentages to match the survey categories. Sources are indicated below:

- Fact sheet from the Provice of Ontario for Visible Minorities [link](https://www.ontario.ca/document/2016-census-highlights/fact-sheet-9-ethnic-origin-and-visible-minorities): Used to obtain percentages for Arab, Black, East Asian/Pacific Islander (adding Chinese, Korean, and Japanese percentages), Latin American, Mixed (using the percentage for "multiple visible minority"), Other (obtained by adding the Southeast Asian, Filipino, West Asian, and Minority not identified elsewhere percentages), South Asian. _Accessed on January 05, 2022_

- Census Profile for Ontario [link](https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/prof/details/page.cfm?Lang=E&Geo1=PR&Code1=35&Geo2=PR&Code2=01&SearchText=Ontario&SearchType=Begins&SearchPR=01&B1=Aboriginal%20peoples&TABID=1&type=1): Used to obtain percentage of Aboriginal population. _Accessed on January 05, 2022_

- Wikipedia entry for Ontario demographics [link](https://en.wikipedia.org/wiki/Demographics_of_Ontario): To corroborate that the percentage of population reported as "European" in this website matched the percentage obtained for "White Caucasian" that was independently obtained by obtaining the difference in population proportion after subtracting the sum of Visible Minorities and Aboriginal Population percentages.

| Ethnicity/Race | percentage |
|----------------|:-----------|
| Arab           | 1.6%       |
| Black          | 4.7%       |
| East Asian/Pacific Islander     | 6.6%       |
| Indigenous     | 2.8%       |
| Latin American | 1.5%       |
| Mixed          | 1.0%       |
| Other          | 5.2%       |
| South Asian    | 8.7%       |
| White Caucasian| 67.8%      |
| **Total**      | 99.9%      |
: Reference Data from Race/Ethnicity

Visible Minorities: 29.3%
Not a Visible Minority: 70.7% (Aboriginal 2.8%, White Caucasian 67.8%)

@@

The question from the survey was "What is your household annual income?". To compare the results of this answer, we used the "Household total income groups in 2015 for private households" from the Census data (available in the Census Data for Ontario website [link](https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/prof/details/page.cfm?Lang=E&Geo1=PR&Code1=35&Geo2=PR&Code2=01&SearchText=Ontario&SearchType=Begins&SearchPR=01&B1=Income&TABID=1&type=1)).

| Household income range (CAD) | Percentage |
|------------------------------|:-----------|
| \< 15,000                    | 5.7%       |
| 15,000 - 24,999              | 7.5%       |
| 25,000 - 39,999              | 11.6%      |
| 40,000- 59,999               | 15.4%      |
| 60,000 - 89,999              | 19.5%      |
| \>90,000                     | 40.3%      |

@@

The brackets for income in the census data are different than the brackets used in the survey. The census does CAD 4,999 brackets (e.g., CAD 5,000- CAD 9,9999) up to CAD 49,999, followed by CAD 9,999 brackets up to CAD 99,999. After that, the brackets increase to CAD 24,999, and therefore, one cannot obtain percentages for the 90,000-109,999 and \>110,000 brackets from the survey.

Therefore, an additional category for income will be created in the dataset so it matches the information from the Census.

@@

clean_data<-clean_data%>%
    mutate(income_ord = case_when(
                                  income %in% "90000_109999" ~ "over_90000",
                                  income %in%"over_110000" ~ "over_90000",
                                  TRUE ~ as.character(income)
                                ))%>%
  mutate(income_ord=as.factor(income_ord))

p3<-clean_data%>%
  mutate(income_ord = fct_relevel(income_ord,
                                  "under_15000",
                                  "15000_24999",
                                  "25000_39999",
                                  "40000_59999",
                                  "60000_89999",
                                  "over_90000"
                                ))

ggplot(p3%>% count(income_ord)%>%  # count number in each group
         mutate(pct=n/sum(n),               # Calculate percent within each group
                ypos = 200),  # Calculate label positions
       aes(income_ord, n, fill=income_ord)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct*100),"%"), y=ypos))+
  theme_classic()+
  thm2

@@

Next, we are interested in analyzing the geographical distribution of the answers the participants within the geographical areas defined by the Association of Municipalities from Ontario. There are two parts to the geographical analysis:

1) Add to the dataset the data for the municipalities so each city has the corresponding region it belongs to.
1) Determine percentage of answers that correspond to Toronto. This is important because although the Census regions do not match the exact geographical boundaries of the municipalities and thus a direct correction for this cannot be easily obtained, Toronto is the largest urban area in the province and it is important to identify if corrections are needed with regard to the proportions from this place in the survey.

### Joining the municipalities data

We have obtained and cleaned the data of the municipalities from the province of Ontario website, and cleaned the dataset to obtain the city names and geographical locations. Further details can be found in `municipalities.qmd` in the `data_cleaning` directory. Next, we will join and match the geographical location from the municipalities dataset to the clean dataset we have obtained in this document so far after removing the entries that do not have a corresponding geographical location.

@@

### Merging missing municipalities and geographic area names

@@

## Proportion of answers within Toronto

The final step is of the geographical analysis is to determine the proportion of answers from Toronto.

clean_data<-clean_data%>%
  mutate(location=as.character(case_when(city=="Toronto"~"Toronto",
                            city!="Toronto"~ "All other cities"
                          )))%>%
  mutate(location=as.factor(location))

ggplot(clean_data %>% count(location) %>%    # then count number in each group
         mutate(pct=n/sum(n),               # Calculate percent within each group
                ypos = 200),  # Calculate label positions
       aes(location, n, fill=location)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct*100),"%"), y=ypos))+
  theme_classic()+
  thm1

@@

The proportion of responses from the city of Toronto is 31.3% of the total in the clean dataset. From the (Census data)[https://www.toronto.ca/wp-content/uploads/2022/02/92e3-City-Planning-2021-Census-Backgrounder-Population-Dwellings-Backgrounder.pdf], the city of Toronto as of 2021 has a population of 2,794,356, corresponding to 19.6% of the total population of the Province.

At this point, the data has been cleaned and prepared for formal analysis.

The clean dataset will be saved as a *.csv file.

```{r,save-clean-dataset}
#| echo: false

#write.csv(clean_data,here("data","clean_dataset.csv"),row.names = FALSE)
```
